EGGs are adhesive!\\ An algebraic account of equality saturation

### 2025-09-09 jlamp/jlamp.tex
minima

Having proved stability of $\pbc$-pushouts, we now turn to prove that they are Van Kampen with respect to regular monos.
%
A proof is found in \Cref{proof:pbvk}.

@@

\begin{rem}\label{rem:obv1}
By definition, there exist functors $O_{\eq}$, $R_{\eq}$ and $W_{\eq}$ as in the diagrams below, where the horizontal arrows are full and faithful inclusions.
\[\xymatrix{\Eqlmo \ar[r]^-{M_{\mathsf{eq}}} \ar[d]_{O_{\mathsf{eq}}}& \EqHyp\ar[d]^{T} & \EqsHyp \ar[r]^-{S_{\mathsf{eq}}} \ar[d]_{R_{\mathsf{eq}}} & \EqHyp \ar[d]^{T} & \EqpsHyp \ar[r]^-{S_{p, \mathsf{eq}}} \ar[d]_{W_{\mathsf{eq}}} & \EqHyp\ar[d]^{T}\\ \lmo \ar[r]_{M} & \hyp & \shyp \ar[r]_{S} & \hyp  & \pshyp \ar[r]_-{S_p}& \hyp}\]

@@

\begin{restatable}{lem}{equ}\label{lem:equ} The following are true:
	\begin{enumerate}
		\item 	$\Eqlmo$ is closed in $\EqHyp$ under equalizers, binary products and pullbacks.
		\item $\EqsHyp$ is closed in $\EqHyp$ under connected limits; 
		\item $\EqpsHyp$ is closed in $\EqHyp$ under pullbacks.
	\end{enumerate}
\end{restatable}

@@

\begin{restatable}{lem}{coequ}\label{lem:coequ}
 The following are true:
\begin{enumerate}
	\item 	$\Eqlmo$ is closed in $\EqHyp$ under pushouts along morphism whose image through $T$ is downward-closed.
	\item $\EqpsHyp$ is closed in $\EqHyp$ under colimits.
\end{enumerate}
\end{restatable}

@@

\begin{defi}We define the following two classes of arrows:
\begin{itemize}
	\item $\pbl$ is the class of all arrows in $\Eqlmo$ sent by $M_{\eq}$ to $\pbc$;
	\item $\pbs$ is the class of all arrows in $\EqpsHyp$ sent by $S_{p,\eq}$ to $\pbc$.
\end{itemize}
\end{defi}

@@

We are now going to generalize the work done in the previous section equipping term graphs with equivalence relation. First of all we need a notion of labelling for $\EqHyp$.

\begin{defi}
	Let $\mathcal{G}$ be an hypergraph, a \emph{labelled hypergraph with equivalence} is a pair $(\mathcal{H}, (f,g))$ where $\mathcal{H}$ is an object of $\EqHyp$ and $(f,g)$ a morphism $T(\mathcal{H})\to \mathcal{G}$ of $\hyp$. A \emph{morphism of labelled hypergraphs with equivalence} between $(\mathcal{H}, (f,g))$ and $(\mathcal{H}', (h,k))$ is an arrow $(a,b, c)\colon \mathcal{H}\to \mathcal{H}'$ such that:
	\[h\circ a = f \qquad k\circ b = g \].

	We denote the resulting category by $\EqHyps$.
\end{defi}

@@

Let $V_\mathcal{G}\colon \EqHyps\to \EqHyp$ be the functor forgetting the labelling and $\pbc_\mathcal{G}$ the class of morphisms in $\EqHyps$ whose image in $\EqHyp$ is in $\pbc$. From \Cref{prop:eqhyp_complete,cor:pbapp}, the second point of \Cref{thm:slice-functors} and from \Cref{cor:eqade}, we can deduce the following.

@@

\begin{defi}Let$\mathcal{G}$ be an hypergraph and $(\mathcal{H},(f,g))$ labelled hypergraph with equivalence.  We say that $(\mathcal{H},(f,g))$ is a \emph{left-monogamous} (resp.~\emph{separated, pruned}) $\mathcal{G}$-hypergraph if $\mathcal{H}$ is left-monogamous (resp.~separated, pruned).

@@

 \begin{prop}
 	Let $M_{\eq}\colon \Eqlmo\to \EqHyp$ and $S_{\eq}\colon \EqsHyp \to \EqHyp$ be the inclusion functors. Let, moreover, $\Gamma_{R(\mathcal{G})}\colon 1\to \EqHyp$ be the constant functor in $R(\mathcal{G})$. Then
 	\begin{enumerate}
 		\item $\lmo_{\eq, \mathcal{G}}$ is equivalent to $\comma{M_\eq}{\Gamma_{R(\mathcal{G})}}$;
 		\item $\shyp_{\eq, \mathcal{G}}$ is equivalent to $\comma{S_\eq}{\Gamma_{R(\mathcal{G})}}$; 
 		\item $\pshyp_{\mathcal{G}}$ is equivalent to $\pshyp_{\eq}/R(\sp(\pr(\mathcal{G})))$. 
 	\end{enumerate}
 \end{prop}

\begin{defi}We define the following two classes of arrows:
	\begin{itemize}
		\item $\pbl_{\mathcal{G}}$ is the class of all arrows in $\lmo_{\eq, \mathcal{G}}$ sent by $M_{\eq, \mathcal{G}}$ to $\pbc_{\mathcal{G}}$;
		\item $\pbs_{\mathcal{G}}$ is the class of all arrows in $\pshyp_{\eq, \mathcal{G}}$ sent by the inclusion  to $\pbc_{\mathcal{G}}$.
	\end{itemize}
\end{defi}

>>>>

Having proved the stability of $\pbc$-pushouts, we now turn to prove that they are Van Kampen with respect to regular monos.
%
A proof is found in \Cref{proof:pbvk}.

@@

\begin{rem}\label{rem:obv1}
By definition, there exist functors $O_{\eq}$, $R_{\eq}$ and $W_{\eq}$ as in the diagrams below, where the horizontal arrows are full and faithful inclusions
\[\xymatrix{\Eqlmo \ar[r]^-{M_{\mathsf{eq}}} \ar[d]_{O_{\mathsf{eq}}}& \EqHyp\ar[d]^{T} & \EqsHyp \ar[r]^-{S_{\mathsf{eq}}} \ar[d]_{R_{\mathsf{eq}}} & \EqHyp \ar[d]^{T} & \EqpsHyp \ar[r]^-{S_{p, \mathsf{eq}}} \ar[d]_{W_{\mathsf{eq}}} & \EqHyp\ar[d]^{T}\\ \lmo \ar[r]_{M} & \hyp & \shyp \ar[r]_{S} & \hyp  & \pshyp \ar[r]_-{S_p}& \hyp}\]

@@

\begin{restatable}{lem}{equ}\label{lem:equ} The following facts hold
	\begin{enumerate}
		\item 	$\Eqlmo$ is closed in $\EqHyp$ under equalizers, binary products and pullbacks.
		\item $\EqsHyp$ is closed in $\EqHyp$ under connected limits; 
		\item $\EqpsHyp$ is closed in $\EqHyp$ under pullbacks.
	\end{enumerate}
\end{restatable}

@@

\begin{restatable}{lem}{coequ}\label{lem:coequ}
 The following facts hold
\begin{enumerate}
	\item 	$\Eqlmo$ is closed in $\EqHyp$ under pushouts along morphisms whose image through $T$ is downward-closed.
	\item $\EqpsHyp$ is closed in $\EqHyp$ under colimits.
\end{enumerate}
\end{restatable}

@@

\begin{defi}We define two classes of arrows
\begin{itemize}
	\item $\pbl$ is the class of all arrows in $\Eqlmo$ sent by $M_{\eq}$ to $\pbc$;
	\item $\pbs$ is the class of all arrows in $\EqpsHyp$ sent by $S_{p,\eq}$ to $\pbc$.
\end{itemize}
\end{defi}

@@

We are now going to generalize the work done in the previous sections equipping term graphs with an equivalence relation. First of all we need a notion of labelling for $\EqHyp$.

\begin{defi}
	Let $\mathcal{G}$ be an hypergraph, a \emph{labelled hypergraph with equivalence} is a pair $(\mathcal{H}, (f,g))$ where $\mathcal{H}$ is an object of $\EqHyp$ and $(f,g)$ a morphism $T(\mathcal{H})\to \mathcal{G}$ of $\hyp$. A \emph{morphism of labelled hypergraphs with equivalence} between $(\mathcal{H}, (f,g))$ and $(\mathcal{H}', (h,k))$ is an arrow $(a,b, c)\colon \mathcal{H}\to \mathcal{H}'$ such that $h\circ a = f$ and $k\circ b = g$.
%
	We denote the resulting category by $\EqHyps$.
\end{defi}

@@

Let $V_\mathcal{G}\colon \EqHyps\to \EqHyp$ be the functor forgetting the labelling and $\pbc_\mathcal{G}$ the class of morphisms in $\EqHyps$ whose image in $\EqHyp$ is in $\pbc$. From \Cref{prop:eqhyp_complete} and \Cref{cor:pbapp}, the second point of \Cref{thm:slice-functors} and from \Cref{cor:eqade}, we can deduce the following.

@@

\begin{defi}Let$\mathcal{G}$ be an hypergraph and $(\mathcal{H},(f,g))$ a labelled hypergraph with equivalence.  We say that $(\mathcal{H},(f,g))$ is a \emph{left-monogamous} (resp.~\emph{separated, pruned}) $\mathcal{G}$-hypergraph if $\mathcal{H}$ is left-monogamous (resp.~separated, pruned).

@@

 \begin{prop}
 	Let $M_{\eq}\colon \Eqlmo\to \EqHyp$ and $S_{\eq}\colon \EqsHyp \to \EqHyp$ be the inclusion functors. Moreover, let $\Gamma_{R(\mathcal{G})}\colon 1\to \EqHyp$ be the constant functor in $R(\mathcal{G})$. Then
 	\begin{enumerate}
 		\item $\lmo_{\eq, \mathcal{G}}$ is equivalent to $\comma{M_\eq}{\Gamma_{R(\mathcal{G})}}$;
 		\item $\shyp_{\eq, \mathcal{G}}$ is equivalent to $\comma{S_\eq}{\Gamma_{R(\mathcal{G})}}$; 
 		\item $\pshyp_{\mathcal{G}}$ is equivalent to $\pshyp_{\eq}/R(\sp(\pr(\mathcal{G})))$. 
 	\end{enumerate}
 \end{prop}

\begin{defi}We define two classes of arrows
	\begin{itemize}
		\item $\pbl_{\mathcal{G}}$ is the class of all arrows in $\lmo_{\eq, \mathcal{G}}$ sent by $M_{\eq, \mathcal{G}}$ to $\pbc_{\mathcal{G}}$;
		\item $\pbs_{\mathcal{G}}$ is the class of all arrows in $\pshyp_{\eq, \mathcal{G}}$ sent by the inclusion  to $\pbc_{\mathcal{G}}$.
	\end{itemize}
\end{defi}
